---
title: "mlVAR on PCs"
output: html_document
date: "2026-05-25"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(mnet)
library(dplyr)
library(mlVAR)
library(tidyverse)
library(qgraph)
library(stringr)

## Change based on your path
name="v2_day"
### FEATURE == end_depressed_binary !!

brighten_dir = "/Users/klj9278/Library/CloudStorage/Box-Box/Kaley_research_NYU/EMA_projects/smartphone_sensor_modelling_may26/data/interim"
output_dir =  "/Users/klj9278/Library/CloudStorage/Box-Box/Kaley_research_NYU/EMA_projects/smartphone_sensor_modelling_may26/results/charts/mlVAR/trainval"

## estimator = The estimator to be used. "lmer" for sequential univariate multi-level estimation, "Mplus" for multivariate Bayesian estimation (requires Mplus), and "lm" for fixed effects estimation. 

## contemporaneous = How should the contemporaneous networks be estimated? These networks are always estimated post-hoc by investigating the residuals of the temporal models. "correlated" and "orthogonal" run second multi-level models in which the networks are estimated using node-wise estimation. "fixed" and "unique" simply correlate the residuals, either by computing one network for all subjects (fixed) or a single network per per subject. 

## temporal = How should the temporal effects be estimated? "correlated" estimates correlated random effects, "orthogonal" estimates non-correlated random effects and "fixed" estimates a model in which only the intercept is random. Defaults to "correlated" when the number of variables is less than 6 and "orthogonal" otherwise. "unique" uses lm to estimate an unique model for each subject. 

```


# mlVAR Package

---

## Types of Networks 

### Temporal Network
The temporal network represents how variables predict each other across time points (from t to t+1).
* Structure: It is a directed network where edges (arrows) indicate the predictive relationship between a variable at one measurement and another variable (or itself) at the next measurement.
* Statistical Basis: Edges are based on Vector Autoregressive (VAR) coefficients. In an mlVAR framework, these are typically estimated through univariate multilevel regressions where each variable is predicted by the lagged versions of all variables in the network.
* Interpretation: These edges represent carry-over effects or "inertia" (when a node predicts itself over time)

### Contemporaneous Network
* Structure: It is an undirected graphical model, specifically a Pairwise Markov Random Field (PMRF).
* Statistical Basis: Edges represent partial correlations between the residuals (also called innovations) of the VAR model. These are the associations that remain after the temporal effects from the previous time point have been "partialled out" or controlled for.
* Interpretation: This network captures dynamics that occur at faster timescales than the measurement interval (e.g., if measurements are taken every three hours, any interaction happening within minutes will appear here). An edge indicates that two variables are conditionally dependent at the same time point, given all other variables in the system


--- 

## Choosing "Estimator" 
* "lmer" — Sequential Univariate Multi-level (Default)
  * Runs a separate mixed-effects regression (via lme4) for each variable, one at a time ("nodewise" estimation)
  * Models random effects — each person gets their own temporal coefficients, shrunk toward the group mean
  * Temporal and between-subject effects come directly from the models; contemporaneous effects are estimated post-hoc by correlating the residuals across equations
  * The fitted model is technically misspecified with respect to a true joint multilevel VAR, since it uses sample means as proxies for population means and applies a disjoint pseudolikelihood approach — but it's fast, open-source, and the most widely used option Springer
  * Best for: most applied research; good balance of speed, flexibility, and handling of individual differences


* "Mplus" — Multivariate Bayesian Estimation
  * Generates a Mplus model, runs the analysis, and reads results back into R; Mplus 8 is required CRAN
  * Estimates all variables jointly (truly multivariate), not equation-by-equation
  * Uses Bayesian MCMC (with iterations and chains arguments controlling the sampler)
  * Can estimate random effects for contemporaneous relationships, which lmer cannot do properly — it is recommended to set contemporaneous = "fixed", though not required CRAN
  * Does not support the dayvar argument — day structure must be handled manually via beepvar GitHub
Best for: when you need the most statistically rigorous, fully joint estimation and have Mplus licensed


* "lm" — Fixed Effects Only
  * Runs plain OLS regression per variable, with no random effects
  * Ignores between-person variation entirely — estimates one set of coefficients assumed to apply to everyone equally
  * Faster and simpler, but treats all subjects as identical in their dynamics
  * No shrinkage, no person-level variance components
  * Best for: quick exploratory analysis, or when your sample is so small that mixed-effects models won't converge; not recommended for serious inference in ESM/EMA data

---

## Estimating Random Effects 
General Differences Between Estimation Methods:
* **"correlated"**: This method estimates correlated random slopes for within-cluster predictors. Theoretically, this allows for the possibility that an individual's specific deviation from the group average in one dynamic relationship (e.g., the effect of "worry" on "sadness") is statistically related to their deviation in another relationship.
* **"orthogonal"**: This estimates uncuncorrelated (orthogonal) random slopes. While it still allows every individual to have their own unique dynamic network, it assumes that these individual differences are independent of one another across different edges in the network.
* **"fixed"**: This estimates only a random intercept and assumes the dynamic relationships (the edges) are fixed across the entire sample. In this model, individuals can have different average levels of symptoms, but the way symptoms influence each other over time is assumed to be identical for everyone.
* **"default"**: This is a pragmatic setting that chooses "correlated" when the number of variables is small (6 or fewer) and "orthogonal" for larger networks to manage computational demands [Source context].

**Notes**
* *Bias and Misspecification*: The `mlVAR` package uses a stepwise estimation approach (nodewise regressions) rather than fitting a single joint model. This can lead to biases, particularly in the "between-person" network (how individuals' averages relate), which may not be resolved simply by choosing a different random effect structure.
* *Significance Testing*: When performing inference on group differences (e.g., comparing patients to healthy controls), the choice of random effects can impact the sampling distribution. The sources recommend using permutation tests in these scenarios, as they are less sensitive to violations of distributional assumptions compared to parametric tests.

### Choosing `contemporaneous` Option

This controls how the **residual correlation network** (what's correlated *within the measurement observation*, after removing lagged effects between periods) is estimated.

| Option | What it does | Use when |
|---|---|---|
| `"correlated"` | 2nd-stage multilevel model, random effects **allowed to covary** | You have enough data per person and want to model *why* people differ in their contemporaneous structure |
| `"orthogonal"` | 2nd-stage multilevel model, random effects **assumed independent** | Similar to above but faster/more stable; use when correlated version fails to converge |
| `"fixed"` | One single contemporaneous network for **all subjects** | You assume everyone shares the same contemporaneous structure, or small N per person |
| `"unique"` | Separate network **per subject** (just correlates that person's residuals) | Large N per person, you want fully individual networks, don't need group-level inference |

**Key decision:** Do you believe contemporaneous relationships **vary across people** (→ `correlated`/`orthogonal`) or are they **shared** (→ `fixed`)? And do you have enough data per person to support random effects?


### Choosing `temporal` Option

This controls how the **lag-1 regression coefficients** (how past predicts present) are structured across people.

| Option | What it does | Use when |
|---|---|---|
| `"correlated"` | Random slopes, **allowed to covary** with each other | Default for <6 vars; you want the full random effects covariance structure |
| `"orthogonal"` | Random slopes, **assumed independent** | Default for ≥6 vars; more stable with many variables, avoids estimating a huge covariance matrix |
| `"fixed"` | **Only the intercept** is random; slopes are fixed across people | You don't believe temporal dynamics vary by person, or very few observations per person |
| `"unique"` | Fits a **separate `lm` per person** | Very large N per person; you want fully individual models with no pooling/shrinkage |

**Key decision axis:**

```
Few obs/person ──────────────────────────── Many obs/person
"fixed"         "orthogonal"   "correlated"   "unique"
← more pooling/shrinkage          less pooling →
```


**Guidelines**

- **Small ESM study** (e.g. 30–50 obs/person, 5–10 people): `temporal = "orthogonal"`, `contemporaneous = "fixed"`
- **Typical ESM study** (50–100 obs/person, 20–50 people, <6 vars): defaults (`temporal = "correlated"`, `contemporaneous = "correlated"`) are fine
- **Many variables (≥6)**: `temporal = "orthogonal"` to avoid the covariance matrix explosion (p² parameters)
- **Large N per person** (200+): `"unique"` becomes viable and gives you fully individual networks
- **Convergence problems**: step down — `correlated` → `orthogonal` → `fixed`

---
  



```{r checking-v1day}

step="pca"
path <- file.path(brighten_dir, paste0(name, "_trainval_", step, "_outcomes.csv"))
df <- read.csv(path)
df <- df[!is.na(df$end_depressed_binary), ] #drop missing values, people who changed over time.
table(df$end_depressed_binary)

df$end_depressed_binary2 <- as.integer(df$end_depressed_binary + 1)
df$end_depressed_binary2[df$end_depressed_binary2 == 3] <- 2

cols <- grep("^pc_", colnames(df), value = TRUE)



# Check values
# Check for infinite values
sapply(df[, cols], function(x) sum(is.infinite(x)))

# Check for NaN
sapply(df[, cols], function(x) sum(is.nan(x)))

# Check for NA
sapply(df[, cols], function(x) sum(is.na(x)))


# Check for zero-variance columns within a person
# If a person never varies on a variable, regression fails
#zero_var = df %>%
 # group_by(num_id) %>%
  #summarise(across(all_of(cols), sd, na.rm = TRUE)) %>%
  #summarise(across(all_of(cols), function(x) sum(x == 0, na.rm = TRUE))) %>%
  #filter(if_all(all_of(cols), function(x) x < 0.5)) %>%
  #pull(num_id)

# Check NA rate per person per variable
perc_missing_mobility = df %>%
  group_by(num_id) %>%
  summarise(
    mob_na = sum(is.na(pc_commute)),
    mob_n  = n(),
    mob_pct = mob_na / mob_n
  ) %>%
  filter(mob_pct > 0.5) %>%
  arrange(desc(mob_pct))

# Check NA rate per person per variable
perc_missing_mobility_r = df %>%
  group_by(num_id) %>%
  summarise(
    mobr_na = sum(is.na(pc_phq2)),
    mobr_n  = n(),
    mobr_pct = mobr_na / mobr_n
  ) %>%
  filter(mobr_pct > 0.5) %>%
  arrange(desc(mobr_pct))


# Check NA rate per person per variable
perc_missing_calls = df %>%
  group_by(num_id) %>%
  summarise(
    uc_na = sum(is.na(pc_active_mobility)),
    uc_n  = n(),
    uc_pct = uc_na / uc_n
  ) %>%
  filter(uc_pct > 0.5) %>%
  arrange(desc(uc_pct))




# Check NA rate per person per variable
# Remove persons with >50% NA on any variable
nonNA_ids <- df %>%
  group_by(num_id) %>%
  summarise(across(all_of(cols), function(x) mean(is.na(x)))) %>%
  filter(if_all(all_of(cols), function(x) x < 0.7)) %>%
  pull(num_id)


#std_results %>%
#  pivot_longer(-num_id, names_to = "column", values_to = "sd") %>%
#  filter(sd == 0 | is.na(sd)) %>%
#  select(num_id, column)

# Find zero-var columns (all std values are 0 or NA)
#zero_var_cols <- std_results %>%
#  select(-num_id) %>%
#  summarise(across(everything(), ~ all(. == 0 | is.na(.)))) %>%
#  pivot_longer(everything()) %>%
#  filter(value == TRUE) %>%
#  pull(name)

# print(zero_var_cols)



# Drop participants with a var >50% missing and report number dropped
df_clean <- df %>% filter(num_id %in% nonNA_ids)
cat(n_distinct(df_clean$num_id), "participants. Removed", n_distinct(df$num_id) - n_distinct(df_clean$num_id), "for having a var with >50% NaN. .\n")

# Drop participants with a var with 0 variation and report number dropped
# df_clean <- df_clean %>% filter(!num_id %in% zero_var_ids)
# cat(n_distinct(df_clean$num_id), "participants. Removed", n_distinct(df$num_id) - n_distinct(df_clean$num_id), "for having zero-var variables.\n")

num_participants = n_distinct(df_clean$num_id)

```


```{r check-cleaned-v1}
# # Simulate what mlVAR does - check if lagging creates problems
# df_clean %>%
#   group_by(num_id) %>%
#   arrange(num_id) %>%
#   mutate(across(all_of(cols), list(lag = ~lag(.)))) %>%
#   summarise(across(ends_with("_lag"), function(x) sum(is.infinite(x) | is.nan(x), na.rm=TRUE))) %>%
#   filter(if_any(-num_id, function(x) x > 0))
# 
# 
# df_clean2 = df_clean %>% 
#   count(num_id) %>% 
#   filter(n <= 2) %>%
#   arrange(n)
# 
# sapply(df_clean[, cols], function(x) range(x, na.rm=TRUE))
# 
# 
# df_clean %>% 
#   count(num_id) %>% 
#   arrange(n) 
# 
# min_obs <- 40  # conservative cutoff
# 
# enough_obs_ids <- df_clean %>%
#   count(num_id) %>%
#   filter(n >= min_obs) %>%
#   pull(num_id)
# 
# cat("Removing", n_distinct(df_clean$num_id) - length(enough_obs_ids), "participants with <", min_obs, "obs\n")
# 
# df_clean <- df_clean %>% filter(num_id %in% enough_obs_ids)
# cat(n_distinct(df_clean$num_id), "participants remaining\n")
# 
# cat("For DF:", name, "STEP:", step, ":\n")
# saveRDS(l_outlist, file=file.path(brighten_dir, paste0("mnet_output_end_depressed_binary2_lme","_pca", "_", name,".RDS")))

```


```{r groupdiff}
clin <- mlVAR_GC(data = df_clean,  
                        vars = cols, 
                        idvar = "num_id", 
                        groups = "end_depressed_binary2", 
                        test = "permutation", 
                        paired = FALSE, 
                        contemporaneous = "orthogonal", 
                        temporal = "orthogonal", 
                        nP = 1000, nCores = 12, pbar = TRUE)



l_outlist <- list()
l_outlist$GC <- clin

saveRDS(l_outlist, file=file.path(brighten_dir, paste0("mnet_output_end_depressed_binary_g1g2_PCA_", name,".RDS")))



```

```{r indiv-groups}

clin_g1 <- mlVAR(data = df_clean[df_clean$end_depressed_binary2==1,],
                vars = cols,
                estimator = "lmer",
                idvar = "num_id",
                contemporaneous = "orthogonal",
                temporal = "orthogonal")

l_outlist$g1 <- clin_g1


clin_g2 <- mlVAR(data = df_clean[df_clean$end_depressed_binary2==2,],
                vars = cols,
                estimator = "lmer",
                idvar = "num_id",
                contemporaneous = "orthogonal",
                temporal = "orthogonal")

l_outlist$g2 <- clin_g2

saveRDS(l_outlist, file=file.path(brighten_dir, paste0("mnet_output_end_depressed_binary_PCA_g1g2_", name,".RDS")))

```


```{r plot1}


# Plot no-clinical-level at baseline
# P-values for the five parameter types:
pvalues = clin$Pval

# The observed group differences (i.e., the test statistics)
# Can be found in:
empdiff = clin$EmpDiffs

# Specifically, the difference is: group 1 - group 2
# For example:
groupdiff = clin$EmpDiffs$Phi_mean[1,2,]
# The true group difference was -0.4



```



```{r plot2}
# --- Check Results ---

output <- l_outlist$GC
out_g1 <- l_outlist$g1
out_g2 <- l_outlist$g2

# output$Runtime_min
# output$Pval$Lagged_fixed
#
# mean(output$Pval$Lagged_fixed<=0.05)
# sum(output$Pval$Lagged_fixed<=0.05)
#
# output$EmpDiffs

# For Tutorial
l_nets <- list()
l_nets$phi_group0 <- out_g1$results$Beta$mean[, , 1]
l_nets$phi_group1 <- out_g2$results$Beta$mean[, , 1]
l_nets$phi_diff <- output$EmpDiffs$Lagged_fixed[, , 1]
l_nets$phi_diffs_sig <- l_nets$phi_diff
l_nets$phi_diffs_sig[output$Pval$Lagged_fixed>0.05] <- 0

output$Pval$Lagged_fixed

titles <- c("Not Depresseed at 6 Weeks", "Depressed at 6 Weeks",
            "Differences: Group Ending Not Depressed - Group Ending Depressed", "Significant Differences")

# --- Make Figure ---

pdf(file.path(output_dir, paste0("Figure_end_depressed_binary_", step, "_", name, "_", num_participants, "p_.pdf")), width = 18, height = 14)

par(mfrow=c(2,2))


full_labels <- cols
short_labels <- str_remove(cols, 'pc_')



for(i in 1:4) qgraph(t(l_nets[[i]]),
                     layout = "circle",
                     edge.labels = TRUE,
                     title = titles[i],
                     theme = "colorblind",
                     maximum = 0.2,
                     mar = rep(8, 4),      # large mar gives room for long labels
                     labels = short_labels,  # short but readable
                     label.scale = FALSE,
                     label.cex = 1,
                     vsize = 15,           # smaller nodes
                     esize = 10, 
                     asize = 8, 
                     label.prop = 1.5,    # pushes label outside the node boundary
                     edge.label.cex = 0.7,
                     edge.label.bg = TRUE,
                     label.norm = "OOOOOOOOOOOOOOOO")  # reserves space for long labels

dev.off()


```
