---
title: "mlVAR on PCs"
output: html_document
date: "2026-05-25"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(mlVAR)

## Change based on your path
brighten_dir = "/Users/klj9278/Library/CloudStorage/Box-Box/Kaley_research_NYU/EMA_projects/smartphone_sensor_modelling_may26/data/interim"
output_dir =  "/Users/klj9278/Library/CloudStorage/Box-Box/Kaley_research_NYU/EMA_projects/smartphone_sensor_modelling_may26/results/charts/mlVAR/trainval"

## estimator = The estimator to be used. "lmer" for sequential univariate multi-level estimation, "Mplus" for multivariate Bayesian estimation (requires Mplus), and "lm" for fixed effects estimation. 

## contemporaneous = How should the contemporaneous networks be estimated? These networks are always estimated post-hoc by investigating the residuals of the temporal models. "correlated" and "orthogonal" run second multi-level models in which the networks are estimated using node-wise estimation. "fixed" and "unique" simply correlate the residuals, either by computing one network for all subjects (fixed) or a single network per per subject. 

## temporal = How should the temporal effects be estimated? "correlated" estimates correlated random effects, "orthogonal" estimates non-correlated random effects and "fixed" estimates a model in which only the intercept is random. Defaults to "correlated" when the number of variables is less than 6 and "orthogonal" otherwise. "unique" uses lm to estimate an unique model for each subject. 

```


# mlVAR Package

---

## Types of Networks 

### Temporal Network
The temporal network represents how variables predict each other across time points (from t to t+1).
* Structure: It is a directed network where edges (arrows) indicate the predictive relationship between a variable at one measurement and another variable (or itself) at the next measurement.
* Statistical Basis: Edges are based on Vector Autoregressive (VAR) coefficients. In an mlVAR framework, these are typically estimated through univariate multilevel regressions where each variable is predicted by the lagged versions of all variables in the network.
* Interpretation: These edges represent carry-over effects or "inertia" (when a node predicts itself over time)

### Contemporaneous Network
* Structure: It is an undirected graphical model, specifically a Pairwise Markov Random Field (PMRF).
* Statistical Basis: Edges represent partial correlations between the residuals (also called innovations) of the VAR model. These are the associations that remain after the temporal effects from the previous time point have been "partialled out" or controlled for.
* Interpretation: This network captures dynamics that occur at faster timescales than the measurement interval (e.g., if measurements are taken every three hours, any interaction happening within minutes will appear here). An edge indicates that two variables are conditionally dependent at the same time point, given all other variables in the system


--- 

## Choosing "Estimator" 
* "lmer" — Sequential Univariate Multi-level (Default)
  * Runs a separate mixed-effects regression (via lme4) for each variable, one at a time ("nodewise" estimation)
  * Models random effects — each person gets their own temporal coefficients, shrunk toward the group mean
  * Temporal and between-subject effects come directly from the models; contemporaneous effects are estimated post-hoc by correlating the residuals across equations
  * The fitted model is technically misspecified with respect to a true joint multilevel VAR, since it uses sample means as proxies for population means and applies a disjoint pseudolikelihood approach — but it's fast, open-source, and the most widely used option Springer
  * Best for: most applied research; good balance of speed, flexibility, and handling of individual differences


* "Mplus" — Multivariate Bayesian Estimation
  * Generates a Mplus model, runs the analysis, and reads results back into R; Mplus 8 is required CRAN
  * Estimates all variables jointly (truly multivariate), not equation-by-equation
  * Uses Bayesian MCMC (with iterations and chains arguments controlling the sampler)
  * Can estimate random effects for contemporaneous relationships, which lmer cannot do properly — it is recommended to set contemporaneous = "fixed", though not required CRAN
  * Does not support the dayvar argument — day structure must be handled manually via beepvar GitHub
Best for: when you need the most statistically rigorous, fully joint estimation and have Mplus licensed


* "lm" — Fixed Effects Only
  * Runs plain OLS regression per variable, with no random effects
  * Ignores between-person variation entirely — estimates one set of coefficients assumed to apply to everyone equally
  * Faster and simpler, but treats all subjects as identical in their dynamics
  * No shrinkage, no person-level variance components
  * Best for: quick exploratory analysis, or when your sample is so small that mixed-effects models won't converge; not recommended for serious inference in ESM/EMA data

---

## Estimating Random Effects 
General Differences Between Estimation Methods:
* **"correlated"**: This method estimates correlated random slopes for within-cluster predictors. Theoretically, this allows for the possibility that an individual's specific deviation from the group average in one dynamic relationship (e.g., the effect of "worry" on "sadness") is statistically related to their deviation in another relationship.
* **"orthogonal"**: This estimates uncuncorrelated (orthogonal) random slopes. While it still allows every individual to have their own unique dynamic network, it assumes that these individual differences are independent of one another across different edges in the network.
* **"fixed"**: This estimates only a random intercept and assumes the dynamic relationships (the edges) are fixed across the entire sample. In this model, individuals can have different average levels of symptoms, but the way symptoms influence each other over time is assumed to be identical for everyone.
* **"default"**: This is a pragmatic setting that chooses "correlated" when the number of variables is small (6 or fewer) and "orthogonal" for larger networks to manage computational demands [Source context].

**Notes**
* *Bias and Misspecification*: The `mlVAR` package uses a stepwise estimation approach (nodewise regressions) rather than fitting a single joint model. This can lead to biases, particularly in the "between-person" network (how individuals' averages relate), which may not be resolved simply by choosing a different random effect structure.
* *Significance Testing*: When performing inference on group differences (e.g., comparing patients to healthy controls), the choice of random effects can impact the sampling distribution. The sources recommend using permutation tests in these scenarios, as they are less sensitive to violations of distributional assumptions compared to parametric tests.

### Choosing `contemporaneous` Option

This controls how the **residual correlation network** (what's correlated *within the measurement observation*, after removing lagged effects between periods) is estimated.

| Option | What it does | Use when |
|---|---|---|
| `"correlated"` | 2nd-stage multilevel model, random effects **allowed to covary** | You have enough data per person and want to model *why* people differ in their contemporaneous structure |
| `"orthogonal"` | 2nd-stage multilevel model, random effects **assumed independent** | Similar to above but faster/more stable; use when correlated version fails to converge |
| `"fixed"` | One single contemporaneous network for **all subjects** | You assume everyone shares the same contemporaneous structure, or small N per person |
| `"unique"` | Separate network **per subject** (just correlates that person's residuals) | Large N per person, you want fully individual networks, don't need group-level inference |

**Key decision:** Do you believe contemporaneous relationships **vary across people** (→ `correlated`/`orthogonal`) or are they **shared** (→ `fixed`)? And do you have enough data per person to support random effects?


### Choosing `temporal` Option

This controls how the **lag-1 regression coefficients** (how past predicts present) are structured across people.

| Option | What it does | Use when |
|---|---|---|
| `"correlated"` | Random slopes, **allowed to covary** with each other | Default for <6 vars; you want the full random effects covariance structure |
| `"orthogonal"` | Random slopes, **assumed independent** | Default for ≥6 vars; more stable with many variables, avoids estimating a huge covariance matrix |
| `"fixed"` | **Only the intercept** is random; slopes are fixed across people | You don't believe temporal dynamics vary by person, or very few observations per person |
| `"unique"` | Fits a **separate `lm` per person** | Very large N per person; you want fully individual models with no pooling/shrinkage |

**Key decision axis:**

```
Few obs/person ──────────────────────────── Many obs/person
"fixed"         "orthogonal"   "correlated"   "unique"
← more pooling/shrinkage          less pooling →
```


**Practical guidance**

- **Small ESM study** (e.g. 30–50 obs/person, 5–10 people): `temporal = "orthogonal"`, `contemporaneous = "fixed"`
- **Typical ESM study** (50–100 obs/person, 20–50 people, <6 vars): defaults (`temporal = "correlated"`, `contemporaneous = "correlated"`) are fine
- **Many variables (≥6)**: `temporal = "orthogonal"` to avoid the covariance matrix explosion (p² parameters)
- **Large N per person** (200+): `"unique"` becomes viable and gives you fully individual networks
- **Convergence problems**: step down — `correlated` → `orthogonal` → `fixed`

---


```{r}
model_file <- function(name) {
  path <- file.path(brighten_dir, paste0(name, "_trainval_pca.csv"))
  df <- read.csv(path)
  pc_cols <- grep("^pc_", colnames(df), value = TRUE)
  
  
  model <- mlVAR(df, vars = pc_cols, idvar = "num_id", lags = 1,
                 estimator = "lmer",
                 contemporaneous = "orthogonal", temporal = "orthogonal")
  
  return(model)

}

save_plots <- function(name, model) {
  # Save fixed effects plots
  res <- residuals(model)  # returns a data frame, one column per variable
  write.csv(res, file.path(output_dir, paste0(name, "_residuals.csv")))

  for (type in c("temporal", "contemporaneous", "between")) {
    
    png(file.path(output_dir, paste0(name, "_", type, ".png")), 
        width = 800, height = 800)
    plot(model, type = type, lag = 1)
    
    # Get all three network types as matrices
    temporal_mean   <- model$results$Beta$mean        # group mean
    write.csv(temporal_mean, file.path(output_dir, paste0(name, "_", type, "_mean.csv")))
    
    temporal_subject  <- model$results$Beta$subject     # per-person, array of matrices
    write.csv(temporal_subject, file.path(output_dir, 'trainval', paste0(name, "_", type, "_subject.csv")))
  }
}

v1_day = model_file('v1_day')
v2_day = model_file('v2_day')
v1_week = model_file('v1_week')
v2_week = model_file('v2_week')

#save_plots('v1_day', v1_day)
#save_plots('v2_day', v2_day)
#save_plots('v1_week', v1_week)
#save_plots('v2_week', v2_week)

```

# Group Difference Testing
* **FPR Inflation**: For fixed lagged effects, the observed FPR was often higher than the specified α (e.g., a test set at α=0.05 might yield an FPR of 0.07). Haslbeck attributed this to misspecification inherent in the two-step estimation approach of mlVAR.
* **Increased Variance**: The study found that as real group differences in some parameters grew larger, the spread (variance) of the sampling distribution for other parameters (where no difference existed) also increased. Because the test does not account for this increased spread, it rejects the null hypothesis too often, leading to an FPR that is higher than the α threshold.
* Alignment can also fail in the **opposite direction** (FPR deflation). For random effects variances near zero, the mlVAR algorithm often biases estimates away from zero. This creates a sampling distribution with an upwardly biased variance, making it nearly impossible to find a significant result at standard α levels, resulting in an observed FPR of 0 regardless of whether α is set to 0.01 or 0.10.

```{r all-res}
library(moments)

# Testing assumptions of the parametric and permutation tests

for (model in list(v1_day, v2_day, v1_week, v2_week)) {

  res <- read.csv(file.path(output_dir, paste0(name, "_residuals.csv")))
  pc_cols <- grep("^pc_", colnames(df), value = TRUE)


  # PARAMETRIC VS PERMUTATION TESTS
  # Visual checks
  par(mfrow = c(2, length(pc_cols)))
  for (v in pc_cols) {
    hist(res[[v]], main = paste("Residuals:", v), breaks = 30)
    qqnorm(res[[v]], main = v); qqline(res[[v]])
  }
  
  # Formal test (Shapiro-Wilk, sensitive to large N — use with caution)
  for (v in pc_cols) {
    cat(v, ":", shapiro.test(sample(na.omit(res[[v]]), min(5000, sum(!is.na(res[[v]])))))$p.value, "\n")
  }
  
  # Skewness and kurtosis
  for (v in pc_cols) {
    cat(v, "- skew:", skewness(na.omit(res[[v]])), 
           "  kurt:", kurtosis(na.omit(res[[v]])), "\n")
  }
  

  for (i in seq_along(pc_cols)) {
    mod <- model$output[[1]][[pc_cols[i]]]
    
    if (is.null(mod)) {
      cat("\n--- Variable:", pc_cols[i], "--- SKIPPED (NULL model)\n")
      next
    }

    re <- lme4::ranef(model$output[[1]][[pc_cols[i]]])$num_id
    
    cat("\n--- Variable:", pc_cols[i], "---\n")
    for (j in seq_len(ncol(re))) {
      vals <- re[[j]]
      if (length(unique(vals)) < 3) {
        cat(colnames(re)[j], ": skipped (no variance)\n")
        next
      }
      sw <- shapiro.test(vals)
      cat(colnames(re)[j], ": W =", round(sw$statistic, 3),
          " p =", round(sw$p.value, 3), "\n")
    }
  }
}

  


# Using permutation test

```
### Permutation Test
Using permutation test because the S-W test and qqnorm plots and kurtosis show that the residuals are not normally distributed. 

```{r permutation-test}
library(mnet)

name = 'v1_day'
path <- file.path(brighten_dir, paste0(name, "_trainval_pca.csv"))
df <- read.csv(path)
pc_cols <- grep("^pc_", colnames(df), value = TRUE)
cat(pc_cols)

df$gender <- df$gender_0.0 + 1

v1_day_gender <- mlVAR_GC(data = df,  vars = pc_cols, idvar = "num_id", groups = "gender", test = "permutation", paired = FALSE, contemporaneous = "orthogonal", temporal = "orthogonal", nP = 1000, nCores = 12, pbar = TRUE)



```

```{r save}
l_outlist <- list()
l_outlist$GC <- output

```

